Hierarchical time series forecasting in Emergency Medical Services

### 2022-11-16 manuscript/03-experiment.qmd
Added alternative dataviz graph

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, required by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally freezed for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].

@@

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended incidents (ambulance demand)"
#| fig-subcap:
#|   - "Hierarchical structure: Attended incidents in all country is disaggregated into 3 control areas /regions and then into 7 different healthboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attended incident (i.e. there are 35 different nature of incidents including chest pain , breathing problems, heart attack, stroke, and so on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

@@

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "70%"
#| fig.align: center
#| fig-cap: "Time series features of attended incidents across all levels (1530 series)"
incident_gthf <- readr::read_rds(paste0(storage_folder, "incidents_gt.rds"))
incident_gthf %>%
  features(incident, feat_stl) %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(alpha = 0.5) +
  ggthemes::theme_few() +
  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
```

In addition to displaying the trend and seasonality features, we also visualize few time series at various levels of the aggregation. Figure @fig-dataviz reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

```{r}
#| label: fig-dataviz
#| cache: true
#| dependson: "fig-feature"
#| out.width: "70%"
#| fig-cap: "Time series of attended incidents at various levels."
#| fig-subcap:
#|   - "Whole country"
#|   - "By control area"
#|   - "By health board and control area"
#|   - "By health board and priority"
#|   - "By nature of incident and health board"
#| layout: [[44,-2, 44], [44,-2, 44], [100]]
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot() +
  labs(x = "Date", y = "Incidents") +
  ggthemes::theme_few()

@@

>>>>

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, required by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally frozen for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].

@@

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended incidents (ambulance demand)."
#| fig-subcap:
#|   - "Hierarchical structure: Attended incidents in all country is disaggregated into 3 control areas /regions and then into 7 different healthboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attended incident (i.e. there are 35 different nature of incidents including chest pain , breathing problems, heart attack, stroke, and so on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

@@

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "70%"
#| fig.align: center
#| fig-cap: "Time series features of attended incidents across all levels (1530 series)"
incident_gthf <- readr::read_rds(here::here("data/incidents_gt.rds"))
incident_gthf %>%
  features(incident, feat_stl) %>%
  ggplot(aes(x = trend_strength, y = seasonal_strength_week)) +
  geom_point(alpha = 0.25) +
  labs(x = "Strength of trend", y = "Strength of weekly seasonality")
```

In addition to displaying the trend and seasonality features, we also visualize few time series at various levels of the aggregation. Figure @fig-dataviz2 reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

```{r}
#| label: fig-dataviz
#| eval: false
#| include: false
#| dependson: "fig-feature"
#| out.width: "70%"
#| fig-cap: "Time series of attended incidents at various levels."
#| fig-subcap:
#|   - "Whole country"
#|   - "By control area"
#|   - "By health board and control area"
#|   - "By health board and priority"
#|   - "By nature of incident and health board"
#| layout: [[44,-2, 44], [44,-2, 44], [100]]
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot() +
  labs(x = "Date", y = "Incidents") +
  ggthemes::theme_few()

@@

```{r}
#| label: fig-dataviz2
#| dependson: "fig-feature"
#| out.width: "100%"
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Time series of attended incidents at various levels."
no_x_axis <- theme(axis.title.x=element_blank(),
                   axis.text.x=element_blank(),
                   axis.ticks.x=element_blank())
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot(incident) +
  labs(x = "", y = "Incidents") +
  no_x_axis

p_control <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  group_by(date, region) %>%
  summarise(incident = sum(incident), .groups = "drop") %>%
  ggplot(aes(x = date, y = incident, color = factor(region))) +
  geom_line() +
  labs(y = "Incidents", color = "Control")  +
  no_x_axis

p_board <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  #facet_wrap(vars(factor(region)), scales = "free_y") +
  labs(y = "Incidents", color = "Health board") +
  no_x_axis

p_priority <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) %>%
  mutate(
    category = recode(category, RED = "Red", AMB = "Amber", GRE = "Green"),
    category = factor(category, levels=c("Red","Amber","Green"))
  ) |>
  as_tibble() %>%
  select(-nature, -region) %>%
  ggplot(aes(x = date, y = incident, color = factor(category))) +
  geom_line() +
  scale_color_manual(values = c(Red="#ff3300", Amber = "#E69f00", Green="#009e73")) +
  labs(y = "Incidents", color = "Priority") +
  no_x_axis

selected_nature <- c("CHESTPAIN", "STROKECVA", "BREATHING", "ABDOMINAL")
p_nature <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & !is_aggregated(nature)) %>%
  as_tibble() %>%
  mutate(nature = as.character(nature)) %>%
  filter(nature %in% selected_nature) %>%
  group_by(date, nature, lhb) %>%
  summarise(incident = sum(incident), .groups = "drop") %>%
  ggplot(aes(x = date, y = incident, color = nature)) +
  geom_line() +
  labs(x = "Date", y = "Incidents", color = "Nature of incident")

p_total /
p_control /
p_board /
p_priority /
p_nature
```
